On Adding Pattern Matching to Haskell-based Deeply Embedded Domain Specific Languages

### 2020-12-02 paper/paper.tex
added emails
noindent after every code block

% \renewcommand{\linenumberfont}{\color{blue}}

@@

% TODO: Use these:

@@

\author{David Young\inst{1} \and Mark Grebe\inst{2} \and Andy Gill\inst{1}}
\institute{University of Kansas \and University of Central Missouri}

@@

Haskell EDSLs that support powerful language features like pattern matching
allow these advantages to be brought to problem domains where it is currently
either difficult or impossible to use a language that has these benefits.
However, there is no support for capturing pattern matching builtin to the Haskell
language itself. As a result, EDLs which capture pattern matching must do so
via a plugin in a Haskell compiler such as GHC. For each such EDSL, this mechanism
has needed to be reimplemented each time. Additionally, the task of implementing this
is difficult and error-prone. In this paper, we descibe a \textit{reusable}
framework for capturing pattern matching, as well as lambdas and tail recursion.

@@

Additionally, the following functions mark the parts of the code which the Core
plugin will target:

@@

The \ttt{E} type is the type of expressions in the EDSL.  In this paper, the
data constructors of this type will be described incrementally as they
are needed.

@@

Note the instance definition is automatically generated from the \ttt{Generic}
instance.

@@

\ttt{ERepTy ComplexPair \typeeq (Complex Double, Complex Double)}
\\

Here is an example that demonstrates why this is more useful than if it \textit{fully} deconstructed
each type into \ttt{Either}, \ttt{(,)} and \ttt{()}:

@@

If \ttt{ERepTy ComplexPair} were fully deconstructed into \ttt{((Double, Double),
(Double, Double))}, we would need a \ttt{Num} instance for \ttt{(Double,
Double)}.  What we really want is to use the fact that a \ttt{Num} instance
already exists for \ttt{Complex Double}. This is exactly what preserving this
type information allows us to do. We can later use this preserved type
information in backends, through the corresponding \ttt{Typeable} instances (for
instance, to provide special support for arithmetic on complex numbers).

@@

The above pattern matches could also be written as a single nested pattern
match. Both forms compile down to the same Core representation.

@@

Note that \ttt{ERepTy IntList \typeeq Either () (Int, IntList)}. If \ttt{ERepTy}
attempted to ``fully deconstruct'' \ttt{IntList}, it would send the compiler
into an infinite loop.

@@

\ttt{ERep} and \ttt{ERepTy} provide an interface for transferring values between the EDSL
expression language and the source Haskell language:

@@

The key algebraic instances mentioned before are as follows:

@@

The equality constraint ensures that a canonical type is its own canonical type.

@@

For the moment, we will primarily use this type as a type tag and ignore
the values it can take on.

@@

Note the \ttt{ERepTy a \typeeq a} constraints. This constraint ensures that the type
\ttt{a} is already in canonical form (that is, consists entirely of
\ttt{Either}, \ttt{(,)} and base types).

@@

Now consider at the type of \ttt{abs} when it is specialized to take
\ttt{E (SumMatch a b)} values:

@@

If we postcompose with \ttt{runSumMatch}, we get:

@@

The \ttt{SumMatch a b} value which \ttt{abs} returns is exactly the function
which pattern matches according to its input value.

Likewise for \ttt{ProdMatch a b} values.

@@

To use the new \ttt{f'} function, we repeatedly call it until it gives a
\ttt{Done} value. If it gives a \ttt{Step} value, we pass the the value wrapped
in the \ttt{Step} back into \ttt{f'} and continue.

@@

This technique can be contrasted with the more traditional trampolining
technique for implementing tail recursion. Conventional trampolining uses a
sum type of the result type and a thunk with the code necessary to continue
execution.~\cite{Ganz:99:Trampolined}

@@

The process of transforming a tail recursive function in the DSL to a 
representation which could be passed into a backend and used
to generate a traditional iterative loop goes as follows.

@@

We first transform this to use the \verb|Iter| type. The lambda is brought
inside of the \ttt{internalize (externalize ...)} call and wrapped in a call to
\ttt{runIter}. This inner expression is then eta expanded. Finally, each \verb|case|
branch which does not contain a recursive call is wrapped in a \verb|Done| and
every recursive call is transformed into an application of the \verb|Step| constructor.

@@

After this initial tail recursion transformation, the rest of the plugin brings this
result into the expression language. \verb|TailRec| is substituted for \verb|runIter|,
\verb|DoneExp| is substituted for \verb|Done| and \verb|StepExp| is substituted for \verb|Step|.
Next, the pattern matching transformation occurs, producing the final result:

@@

The \ttt{Typeable} constraints are necessary to lookup correctly typed
values in the variable binding environment later on.

@@

In the Core transformation, each lambda is given a \ttt{Name} with a globally
unique \ttt{Int}, sidestepping any name capture issues.

@@

This type encodes a single variable binding, with values of the form
\ttt{n :=> v}, where \ttt{n} is a typed name and \ttt{v} is the value it is bound to.

@@

Recall that the internal representations of \verb|SumMatch a b| and \verb|ProdMatch a b|
are functions of type \verb|a -> b|. These types are unwrapped by the functions

@@

Given this, the \verb|ProdMatchExp| case of the \verb|prodMatchAbs| function can be
implemented as follows:

@@

See Appendix A for implementation details of \ttt{absEnv}.

@@

For example, \ttt{externalize x} marks the expression \ttt{x}.  If an expression
already has an EDSL type (a type of the form \ttt{E a} for some \ttt{a}), then
the marking procedure ignores it and does not wrap it with \ttt{externalize}
(see $M$ in Figure~\ref{fig:Rewrites}).

>>>>

@@

@@

\author{David Young\inst{1} \and Mark Grebe\inst{2} \and Andy Gill\inst{1}}
\institute{University of Kansas\\\email{$\{$d063y800,andygill$\}$@ku.edu} \and University of Central Missouri\\\email{grebe@ucmo.edu}}

@@

Haskell EDSLs that support powerful language features like pattern matching
allow these advantages to be brought to problem domains where it is currently
either difficult or impossible to use a language that has these benefits.
However, there is no support for capturing pattern matching builtin to the Haskell
language itself. As a result, EDSLs which capture pattern matching must do so
via a plugin in a Haskell compiler such as GHC. For each such EDSL, this mechanism
has needed to be reimplemented each time. Additionally, the task of implementing this
is difficult and error-prone. In this paper, we descibe a \textit{reusable}
framework for capturing pattern matching, as well as lambdas and tail recursion.

@@

\noindent Additionally, the following functions mark the parts of the code which the Core
plugin will target:

@@

\noindent The \ttt{E} type is the type of expressions in the EDSL.  In this paper, the
data constructors of this type will be described incrementally as they
are needed.

@@

\noindent Note the instance definition is automatically generated from the \ttt{Generic}
instance.

@@

\ttt{ERepTy ComplexPair \typeeq (Complex Double, Complex Double)}
\\
\noindent Here is an example that demonstrates why this is more useful than if it \textit{fully} deconstructed
each type into \ttt{Either}, \ttt{(,)} and \ttt{()}:

@@

\noindent If \ttt{ERepTy ComplexPair} were fully deconstructed into \ttt{((Double, Double),
(Double, Double))}, we would need a \ttt{Num} instance for \ttt{(Double,
Double)}.  What we really want is to use the fact that a \ttt{Num} instance
already exists for \ttt{Complex Double}. This is exactly what preserving this
type information allows us to do. We can later use this preserved type
information in backends, through the corresponding \ttt{Typeable} instances (for
instance, to provide special support for arithmetic on complex numbers).

@@

\noindent The above pattern matches could also be written as a single nested pattern
match. Both forms compile down to the same Core representation.

@@

\noindent Note that \ttt{ERepTy IntList \typeeq Either () (Int, IntList)}. If \ttt{ERepTy}
attempted to ``fully deconstruct'' \ttt{IntList}, it would send the compiler
into an infinite loop.

@@

\noindent \ttt{ERep} and \ttt{ERepTy} provide an interface for transferring values between the EDSL
expression language and the source Haskell language:

@@

\noindent The key algebraic instances mentioned before are as follows:

@@

\noindent The equality constraint ensures that a canonical type is its own canonical type.

@@

\noindent For the moment, we will primarily use this type as a type tag and ignore
the values it can take on.

@@

\noindent Note the \ttt{ERepTy a \typeeq a} constraints. This constraint ensures that the type
\ttt{a} is already in canonical form (that is, consists entirely of
\ttt{Either}, \ttt{(,)} and base types).

@@

\noindent Now consider at the type of \ttt{abs} when it is specialized to take
\ttt{E (SumMatch a b)} values:

@@

\noindent If we postcompose with \ttt{runSumMatch}, we get:

@@

\noindent The \ttt{SumMatch a b} value which \ttt{abs} returns is exactly the function
which pattern matches according to its input value. Likewise for \ttt{ProdMatch a b} values.

@@

\noindent To use the new \ttt{f'} function, we repeatedly call it until it gives a
\ttt{Done} value. If it gives a \ttt{Step} value, we pass the the value wrapped
in the \ttt{Step} back into \ttt{f'} and continue.

@@

\noindent This technique can be contrasted with the more traditional trampolining
technique for implementing tail recursion. Conventional trampolining uses a
sum type of the result type and a thunk with the code necessary to continue
execution.~\cite{Ganz:99:Trampolined}

@@

\noindent The process of transforming a tail recursive function in the DSL to a 
representation which could be passed into a backend and used
to generate a traditional iterative loop goes as follows.

@@

\noindent We first transform this to use the \verb|Iter| type. The lambda is brought
inside of the \ttt{internalize (externalize ...)} call and wrapped in a call to
\ttt{runIter}. This inner expression is then eta expanded. Finally, each \verb|case|
branch which does not contain a recursive call is wrapped in a \verb|Done| and
every recursive call is transformed into an application of the \verb|Step| constructor.

@@

\noindent After this initial tail recursion transformation, the rest of the plugin brings this
result into the expression language. \verb|TailRec| is substituted for \verb|runIter|,
\verb|DoneExp| is substituted for \verb|Done| and \verb|StepExp| is substituted for \verb|Step|.
Next, the pattern matching transformation occurs, producing the final result:

@@

\noindent The \ttt{Typeable} constraints are necessary to lookup correctly typed
values in the variable binding environment later on.

@@

\noindent In the Core transformation, each lambda is given a \ttt{Name} with a globally
unique \ttt{Int}, sidestepping any name capture issues.

@@

\noindent This type encodes a single variable binding, with values of the form
\ttt{n :=> v}, where \ttt{n} is a typed name and \ttt{v} is the value it is bound to.

@@

\noindent Recall that the internal representations of \verb|SumMatch a b| and \verb|ProdMatch a b|
are functions of type \verb|a -> b|. These types are unwrapped by the functions

@@

\noindent Given this, the \verb|ProdMatchExp| case of the \verb|prodMatchAbs| function can be
implemented as follows:

@@

\noindent See Appendix A for implementation details of \ttt{absEnv}.

@@

\noindent For example, \ttt{externalize x} marks the expression \ttt{x}.  If an expression
already has an EDSL type (a type of the form \ttt{E a} for some \ttt{a}), then
the marking procedure ignores it and does not wrap it with \ttt{externalize}
(see $M$ in Figure~\ref{fig:Rewrites}).
